PHQ-4-R - Data Cleaning (Study 2)

### 2024-12-06 study2/analysis/1_cleaning.qmd
update data
@AnafNeves

df <- read.csv("../data/data_raw.csv") |> 
   mutate(across(everything(), ~ifelse(.x == "", NA, .x))) 
```

@@

<!-- df[is.na(df$BDI2_Total), c("Participant")] -->

@@

@@

failed <- reject[reject$AttentionCheck_2 == 0 | reject$AttentionCheck_3 < 0.7, "Participant"]
failed <- failed[!is.na(failed)]

reject |> 
  gt::gt() |> 
  gt::data_color(columns = c("AttentionCheck_2", "AttentionCheck_3"), palette=c("red", "green")) |> 
  gt::data_color(columns = "Participant", fn=\(x) ifelse(x %in% failed, "red", "white")) |> 
  gt::opt_interactive(use_compact_mode = TRUE)

df <- filter(df, !Participant %in% failed)
df <- select(df, -starts_with("AttentionCheck"))
```

We dropped `r length(failed)` participants for failing the attention checks.

@@

```{r}
outliers <- df |> 
  select(Participant, BDI2_SD, BDI2_Duration, STAI5_SD, STAI5_Duration) |> 
  standardize(robust=TRUE) |> 
  mutate(SD = rowSums(across(ends_with("SD"))), Duration = rowSums(across(ends_with("Duration"))))

mv_outliers <- df |> 
  select(Participant, 
         matches("BDI2_"), -BDI2_Duration, -BDI2_Order, -BDI2_Total,
         matches("STAI5_"), -STAI5_Duration, -STAI5_Order, -STAI5_General) |> 
  check_outliers(method="mahalanobis_robust", ID="Participant")

df |>
  mutate(Outliers = mv_outliers) |> 
  ggplot(aes(x=STAI5_SD, y=BDI2_SD)) +
  geom_point(aes(color=Sample, shape=mv_outliers), alpha=2/3, size=3) +
  scale_shape_manual(values=c(16, 4)) +
  ggside::geom_xsidedensity(aes(color=Sample), alpha=0.5) +
  ggside::geom_ysidedensity(aes(color=Sample), alpha=0.5) +
  theme_minimal() +
  ggside::theme_ggside_void()

table <- df |> 
  select(Participant, starts_with("AttentionCheck"), BDI2_SD, BDI2_Duration, STAI5_SD, STAI5_Duration) |> 
  mutate_at(vars(ends_with("Duration")), \(x) ifelse(x > 10, 10, x)) |>
  mutate(Participant = fct_reorder(Participant, outliers$Duration)) |> 
  arrange(Participant)

data.frame(Participant = c("Average"), t(sapply(table[2:ncol(table)], mean, na.rm = TRUE))) |>
  rbind(table) |> 
  gt::gt() |> 
  gt::fmt_number() |> 
  # gt::data_color(columns = "Participant", fn=\(x) ifelse(x %in% outliers_half, "orange", "white")) |>
  gt::data_color(columns = ends_with("Duration"), 
                 direction="column", palette = "RdBu") |> 
  gt::data_color(columns = ends_with("SD"), 
                 direction="column", palette = "RdBu", reverse=TRUE) |> 
  gt::data_color(direction="row", rows=1, palette = "grey") |> 
  gt::opt_interactive(use_compact_mode = TRUE)

df <- df[!mv_outliers,]
```

We dropped `r length(mv_outliers)` participants flagged as multivariate outliers using the robust Mahalanobis distance.

@@

```{r}
table1 <- df |> 
  select(Age, 
         PHQ4_Condition, 
         PHQ4 = PHQ4_Total,
         STAI5 = STAI5_General,
         BDI2 = BDI2_Total) |> 
  report::report_sample(group_by = "PHQ4_Condition")

table1$Variable <- str_remove_all(table1$Variable, fixed("Mean "))
table1$Variable <- str_remove_all(table1$Variable, fixed(" (SD)"))

insight::display(table1)
```

```{r}
add_bf <- function(df, var="Age") {
  r <- report::report(t.test(df[[var]] ~ df[["PHQ4_Condition"]]), verbose=FALSE)
  x <- as.data.frame(r)
  f <- as.formula(paste(var, "~ PHQ4_Condition"))
  x$BF <- as.data.frame(BayesFactor::ttestBF(formula=f, data=df))$bf
  x$Parameter <- var
  x
}

rbind(
  add_bf(filter(df, !is.na(Age)), "Age"),
  add_bf(df, var="PHQ4_Total"),
  add_bf(df, "STAI5_General"),
  add_bf(df, "BDI2_Total")
) |> 
  select(Parameter, t, df_error, p, BF) |> 
  mutate(Parameter = str_remove_all(Parameter, fixed("df$"))) |>
  insight::display()
```

@@

```{r}
df |> 
  select(PHQ4_Condition, 
         STAI5 = STAI5_General,
         BDI2 = BDI2_Total,
         PHQ4_Total) |> 
  standardize() |> 
  estimate_density(at = "PHQ4_Condition") |> 
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color = Parameter), linewidth=2) +
  scale_color_manual(values = c("STAI5" = "#E91E63", "BDI2" = "#2196F3", "PHQ4_Total" = "#9C27B0")) +
  labs(x = "Standardized Score", y = "Distribution") +
  facet_wrap(~PHQ4_Condition) +
  theme_bw()
```

@@

```{r}
df$MoodDisorder <- rowSums(select(df, DisorderTreatment_MoodStabilizer, DisorderTreatment_Antidepressant, DisorderTreatment_Anxiolytic))
df$MoodDisorder <- ifelse(df$MoodDisorder > 0, TRUE, FALSE)
df$DisorderHistory <- ifelse(df$DisorderHistory == "Yes", TRUE, FALSE)
```

`r sum(df$MoodDisorder)` participants (`r insight::format_percent(sum(df$MoodDisorder) / nrow(df))`) have a mood disorder, as indexed by the use of a mood stabilizer, antidepressant, or anxiolytic treatment. `r sum(df$DisorderHistory, na.rm=TRUE)` participants (`r insight::format_percent(sum(df$DisorderHistory, na.rm=TRUE) / nrow(df))`) reported having a history of mood disorder.

>>>>

df <- read.csv("../data/data_raw.csv") |> 
  mutate(across(everything(), ~ifelse(.x == "", NA, .x))) |> 
  filter(Age >= 18) |> 
  mutate(Age = ifelse(Age > 90, NA, Age))
```

@@

#### IAS

df <- df |> 
  mutate(IAS_Total = rowSums(select(df, matches("IAS_"), -IAS_Duration, -IAS_Order)))
```

@@

```{r}
outliers <- list()
```

@@

outliers$attentionchecks <- reject[reject$AttentionCheck_2 == 0 | reject$AttentionCheck_3 < 0.95, "Participant"]
outliers$attentionchecks <- outliers$attentionchecks[!is.na(outliers$attentionchecks)]
df$AttentionCheck <- ifelse(df$Participant %in% outliers$attentionchecks, "Failed", "Passed")

reject |> 
  gt::gt() |> 
  gt::data_color(columns = c("AttentionCheck_2", "AttentionCheck_3"), palette=c("red", "green")) |> 
  gt::data_color(columns = "Participant", fn=\(x) ifelse(x %in% outliers$attentionchecks, "red", "white")) |> 
  gt::opt_interactive(use_compact_mode = TRUE)
```

We removed `r length(outliers$attentionchecks)` participants for failing the attention checks.

```{r}
#| eval: false

dat <- mutate(df, Failed = ifelse(Participant %in% outliers$attentionchecks, 1, 0)) |> 
  filter(Gender %in% c("Female", "Male")) |> 
  filter(Experiment_Duration < 30) |> 
  mutate(Gender = as.factor(Gender)) |> 
  select(Failed, Gender, Experiment_Duration, Age, Ethnicity)

m <- glm(Failed ~ Gender / Age, data=dat, family="binomial") 
parameters::parameters(m)

modelbased::estimate_relation(m, length=50) |> 
  ggplot(aes(x=Age, y=Predicted, color=Gender)) +
  geom_line() 
```

### Duration

```{r}
m <- mgcv::gam(Failed ~ s(Experiment_Duration), 
          data=df |> 
            filter(Experiment_Duration < 30) |> 
            mutate(Failed = ifelse(AttentionCheck == "Failed", 1, 0)), 
          family="binomial")

parameters::parameters(m)
plot(estimate_relation(m, length=40))

df |> 
  filter(Experiment_Duration < 30) |> 
  select(Participant, Experiment_Duration, AttentionCheck) |> 
  mutate(Participant = fct_reorder(Participant, Experiment_Duration)) |>
  ggplot(aes(x=Participant, y=Experiment_Duration, fill=AttentionCheck)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

@@

```{r}
#| eval: false

mgcv::gam(Failed ~ s(STAI5_SD), 
          data=mutate(df, Failed = ifelse(AttentionCheck == "Failed", 1, 0)), 
          family="binomial") |>
  parameters::parameters()

df |> 
  select(Participant, STAI5_SD, AttentionCheck) |> 
  mutate(Participant = fct_reorder(Participant, STAI5_SD)) |>
  ggplot(aes(x=Participant, y=STAI5_SD, fill=AttentionCheck)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

### Multivariate Distance

```{r}
dfoutlier <- df |> 
  select(contains("IAS_"), -IAS_Duration, -IAS_Order, -IAS_Total) |>
  # select(contains("BDI2_"), -BDI2_Duration, -BDI2_Order, -BDI2_Total, -BDI2_SD) |> 
  # datawizard::remove_empty_rows() |> 
  performance::check_outliers(
    method=c("optics", "mahalanobis_robust"),
    threshold=list(optics=2.5, optics_xi=0.03)) |> 
  as.data.frame() |> 
  mutate(Participant = fct_reorder(df$Participant, Distance_OPTICS),
         Outlier_AttentionCheck = ifelse(Participant %in% outliers$attentionchecks, 1, 0),
         Outlier_Distance = ifelse(abs(standardize(Distance_OPTICS)) > qnorm(0.995), 1, 0),
         Experiment_Duration = df$Experiment_Duration,
         Outlier = ifelse(Outlier_AttentionCheck == 1, "Failed Attention Checks", "Passed"))

outliers$distance <- as.character(dfoutlier[dfoutlier$Outlier_Distance == 1, "Participant"])
outliers$distance <- outliers$distance[!outliers$distance %in% c(outliers$attentionchecks, outliers$duration)]

m <- mgcv::gam(Outlier_AttentionCheck ~ s(Distance_OPTICS), data = dfoutlier, family = "binomial")
parameters::parameters(m)

estimate_relation(m, length=40) |> 
  ggplot(aes(x=Distance_OPTICS, y=Predicted)) +
  geom_ribbon(aes(ymin=CI_low, ymax=CI_high), alpha=0.2) +
  geom_line() + 
  ggside::geom_xsidedensity(data=dfoutlier, aes(fill=Outlier), alpha=0.3, color=NA)

dfoutlier |> 
  ggplot(aes(x=Participant, y=Distance_OPTICS, fill=Outlier)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

```{r}
df <- filter(df, !Participant %in% c(outliers$attentionchecks, outliers$distance))
df <- select(df, -starts_with("AttentionCheck"))
```

@@

```{r}
df$PHQ4_Duration[df$PHQ4_Duration > 3] <- NA

table1 <- df |> 
  select(Age, 
         PHQ4_Condition, 
         PHQ4_Total,
         PHQ4_Duration,
         STAI5_General,
         BDI2_Total) |> 
  report::report_sample(by = "PHQ4_Condition")

table1$Variable <- str_remove_all(table1$Variable, fixed("Mean "))
table1$Variable <- str_remove_all(table1$Variable, fixed(" (SD)"))
table1$Variable <- c("Age", "PHQ4 (Total)", "PHQ4 (Duration)", "STAI5", "BDI2")
```

```{r}
add_bf <- function(df, var="PHQ4_Duration") {
  dat <- df[, c(var, "PHQ4_Condition")]
  dat <- dat[complete.cases(dat), ]
  r <- report::report(t.test(dat[[var]] ~ dat[["PHQ4_Condition"]]), verbose=FALSE)
  x <- as.data.frame(r)
  f <- as.formula(paste(var, "~ PHQ4_Condition"))
  x$BF <- as.data.frame(BayesFactor::ttestBF(formula=f, data=dat))$bf
  x$Parameter <- var
  x
}

table1_inference <- rbind(
  add_bf(filter(df, !is.na(Age)), "Age"),
  add_bf(df, var="PHQ4_Total"),
  add_bf(df, var="PHQ4_Duration"),
  add_bf(df, "STAI5_General"),
  add_bf(df, "BDI2_Total")
) |> 
  select(Parameter, t, df_error, p, BF) |> 
  mutate(Parameter = str_remove_all(Parameter, fixed("df$")))

table1$BF <- table1_inference$BF
insight::display(table1)
```

@@

```{r}
df |> 
  select(PHQ4_Condition, 
         STAI5 = STAI5_General,
         BDI2 = BDI2_Total,
         PHQ4_Total) |> 
  standardize() |> 
  estimate_density(by = "PHQ4_Condition", method="KernSmooth") |> 
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color = Parameter), linewidth=2) +
  scale_color_manual(values = c("STAI5" = "#E91E63", "BDI2" = "#2196F3", "PHQ4_Total" = "#9C27B0")) +
  labs(x = "Standardized Score", y = "Distribution") +
  facet_wrap(~PHQ4_Condition) +
  theme_bw()
```

@@

```{r}
df$MoodDisorder <- rowSums(select(df, DisorderTreatment_MoodStabilizer, DisorderTreatment_Antidepressant, DisorderTreatment_Anxiolytic))
df$MoodDisorder <- ifelse(df$MoodDisorder > 0, TRUE, FALSE)

df$Depression <- rowSums(select(df, DisorderTreatment_MoodStabilizer, DisorderTreatment_Antidepressant))
df$Depression <- ifelse(df$Depression > 0, TRUE, FALSE)

df$DisorderHistory <- ifelse(df$DisorderHistory == "Yes", TRUE, FALSE)
```

`r sum(df$MoodDisorder)` participants (`r insight::format_percent(sum(df$MoodDisorder) / nrow(df))`) have a mood disorder, as indexed by the use of a mood stabilizer, antidepressant, or anxiolytic treatment. `r sum(df$Depression)` participants (`r insight::format_percent(sum(df$Depression) / nrow(df))`) have a mood disorder, as indexed by the use of a mood stabilizer or antidepressant. `r sum(df$DisorderHistory, na.rm=TRUE)` participants (`r insight::format_percent(sum(df$DisorderHistory, na.rm=TRUE) / nrow(df))`) reported having a history of mood disorder.
